---
title: "Mini-Project 3: Strings and Regular Expressions"
format:
  html: default
---

## Overview

You will find a data set containing string data.  This could be newspaper articles, tweets, songs, plays, movie reviews, or anything else you can imagine.  Then you will answer questions of interest and tell a story about your data using string and regular expression skills you have developed.

Your story must contain the following elements:

- at least 3 str_ functions
- at least 3 regular expressions
- at least 2 illustrative, well-labeled plots or tables
- a description of what insights can be gained from your plots and tables


## Timeline

Mini-Project 3 must be submitted on moodle by **11:00 PM on Mon Apr 15**.    Ideally, you'd add a tab to your quarto webpage with Mini-Project 3, then you can just submit your URL (as long as your code is linked to the page).  You can alternatively submit a knitted pdf showing your code and output.


## Topic Ideas

```{r}
#| include: false
#| message: false

library(lubridate)
library(tidyverse)
```


### Obama tweets

```{r}
#| message: false

#barack <- read_csv("Data/tweets_potus.csv") 
barack <- read_csv("https://proback.github.io/264_fall_2024/Data/tweets_potus.csv")
#michelle <- read_csv("Data/tweets_flotus.csv") 
michelle <- read_csv("https://proback.github.io/264_fall_2024/Data/tweets_flotus.csv")

tweets <- bind_rows(barack %>% 
                      mutate(person = "Barack"),
                    michelle %>% 
                      mutate(person = "Michelle")) %>%
  mutate(timestamp = ymd_hms(timestamp))
```

President Barack Obama became the first US President with an official twitter account, when @POTUS went live on May 18, 2015. (Yes, there was a time before Twitter.)  First Lady Michelle Obama got in on Twitter much earlier, though [her first tweet](https://youtu.be/3Pze1f1x7yU) was not from @FLOTUS. All of the tweets from @POTUS and @FLOTUS are now archived on Twitter as @POTUS44 and @FLOTUS44, and they are available as a csv download from the National Archive.
Read more here: https://obamawhitehouse.archives.gov/blog/2017/01/05/new-lenses-first-social-media-presidency

Potential things to investigate:

- use of specific terms
- use of @, #, RT (retweet), or -mo (personal tweet from Michelle Obama)
- timestamp for date and time trends
- sentiment analysis (after merging in a sentiment library like in Data Science 1)
- anything else that seems interesting!


### Dear Abby advice column

Read in the “Dear Abby” data underlying The Pudding’s [30 Years of American Anxieties article](https://pudding.cool/2018/11/dearabby/).

```{r}
#| message: false

posts <- read_csv("https://raw.githubusercontent.com/the-pudding/data/master/dearabby/raw_da_qs.csv")
```

Take a couple minutes to scroll throgh the 30 Years of American Anxieties article to get ideas for themes that you might want to search for using regular expressions.


### Other sources for string data

- [Other articles from The Pudding](https://github.com/the-pudding/data)
- NY Times headlines from the RTextTools package (see below)
- further analysis with the `bigspotify` data from class
- the options are endless -- be resourceful and creative!

```{r}
#| message: false

library(RTextTools)  # may have to install first
data(NYTimes)
as_tibble(NYTimes)
```

